The glEnd() of Zelda

### 2016-03-15 smeight/paper/paper.tex
First submitted version

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\linewidth]{link3d}
\end{center}\vspace{-0.1in}
\caption{Technical diagram of mathematical equations.} \label{fig:link3d}
% \smallskip
% caption can go here...
% } 
\end{figure}

@@

The NES has a modest CPU that executes game logic. It's a little 8-bit
baby puppy with access to 2 kilobytes of RAM. If you think about it,
the NES's 256x240 pixel screen is already 61,440 pixels, which if it
used the entire 2k of RAM to represent it, we'd have only
$0.2\overline{6}$ bits of RAM per pixel, which makes no sense. At XXX
MhZ, and a generous 1 instruction per pixel, we'd only be able to render
XXX frames per second! This doesn't make sense, and this is because
the CPU does not render the graphics on a NES.

The NES also contains a custom Picture Processing Unit, or PPU, the
8-bit baby puppy's baby ppupy brother. This thing outputs 60 frames of
NTSC or PAL video per second. In fact, since the PPU also drives a CPU
interrupt for each scanline and vblank (XXX true?), and most games use
this for timing,%
%
\footnote{Fun fact: NTSC, the video standard used in the Americas, has
  % XXX how u make ``approx'' symbol
  a \~60Hz frame rate. But PAL, used in Europe and Asia, has a 50Hz
  refresh rate. This means that many games run 16\% slower in Europe
  than the US, so if you grew up playing there, you were playing on
  {\em Easy} mode. {\tt\ ;-)}}
%
in many ways the CPU is subservient to the PPU, kind of like the PPU
is the {\it true} pure heart of the NES. In the context of this paper, you
could say that the CPU is the Ocarina of Time and the PPU is the
Legend of Zelda. It's just an analogy. The PPU is very complicated and
has tetrahedra for hands, like clubs. The CPU communicates with
the PPU not by drawing pixels, but by giving a high-level description
of the screen (by writing it into PPU memory). The chief insight of
this work is that the PPU representation can be interpreted and
rendered in 3D instead of 2D. This allows us to use the exact original
game logic and only change the way it's viewed. Sometimes this even
allows us to see aspects of the game that are actually present but
were not visible on NES hardware. Often it makes the game more
difficult by making things invisible (enemies behind the player or
behind walls) and inducing motion sickness.

@@

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\linewidth]{ppu1} \\
\vspace{1em}
\includegraphics[width=\linewidth]{ppu2}
\end{center}\vspace{-0.1in}
\caption{Pattern tables for the Zelda overworld. In Zelda, the top tiles
are used for sprites, and the bottom are used for the background graphics.
All tiles are 8x8 and use 4 color values (color 0 is transparent). These
color values index into 4 sprite palettes and 4 background palettes, with
some limitations.
} \label{fig:ppu}
\end{figure}

@@

\begin{tabular}{rrl}
    & bit \# & bit deets \\
\hline \\
MSB & 7      & Vertical flip \\
    & 6      & Horizontal flip \\
    & 5      & something weird \\
    & 4      & I don't know \\
    & 3      & unused? who can say \\
    & 2      & bit 2; could be anything \\
LSB & 1,0    & Palette index \\
\end{tabular}

@@

\section{Three to D---$u$ and $v$}\footnote{Spoiler alert because I think
this joke may be too abstruse. ``Free to Be You and Me'' was a
semi-famous kids record (here meaning actual vinyl) by 
% XXX
Milo Thomas. D is ``dimensions.'' $u$ and $v$ are common names of
the axes for texture coordinates in computer graphics.}

@@

\paragraph{Extracting sprites.}
Sprites are the most difficult. Aside from just being more fiddly to
implement, it is tricky to recover an important high-level structure.
Although we know the location of each individual sprite and the two tiles
(Zelda uses ``tall' 8x16 sprites; see Section~\ref{sec:sprites}) that
make it up, most game objects (like enemies) are in fact drawn with
{\it two} sprites. These sprites could be in any of the 64 sprite
slots (it's not at all obvious how Zelda allocates these, or that it
would be similar for other games). Adding to the complexity, sprites
like Link's sword or the pulsing balls of plasma that frogmen shoot
at you from the water {\it just for being an elf} are just a single
sprite, and bosses can be like XXX sprites (XXX add figure). If we
promote sprites to 3D without knowing which ones are supposed to be
part of the same object, then the enemies get cut up along their
sprite seems and it looks super wrong.

@@

\begin{itemize}
\item Create 64 ``pre-sprites'', which are baby sprites. These know
  their coordinates and have the index of a ``parent'' pre-sprite
  (initialized to $-1$).
\item While filling these in, consult a hash-map to see if there are
  already pre-sprites exactly aligned in each of the cardinal
  directions. If so, use a union-find--like algorithm (XXX cite)
  to efficiently update its parent index; each fused sprite has one
  canonical representative whose parent index is $-1$. Add the
  pre-sprite to the hash-map.
\item Another pass pushes the min/max bounds to the canonical representative,
  so that we know how big the fused sprite is.
\item Another pass draws all sprites into the texture data for their
  canonical representative, at the appropriate location. This pass must
  happen in the same order as on the NES hardware, so that overlay effects
  are correct.
\item Finally, create proper sprites to return for each of the fused sprites.
  Copy the used region of the texture data for these to the video card.
\end{itemize}

@@

We also have to face in the same direction as link, to make the
perspective first-person. Link's orientation is at memory location
\verb+0x98+.

Link moves smoothly (with pixel resolution) around the screen, but
turns abruptly. This is sickening to watch. I apply some smoothing
to Link's angle, ...

(This also involves manual steps..)

@@

......

Note that some of the tiles on screen are used to draw a menu at the
top, which is not part of the gameplay area. This still gets interpreted
as

Problem with link always hovering behind you

\section{A Link to the cast} \label{sec:cast}

\nocite{murphy2013first}

>>>>

\begin{figure}[h!t]
\begin{center}
\includegraphics[width=\linewidth]{link3d}
\end{center}\vspace{-0.1in}
\caption{Technical diagram of mathematical equations.} \label{fig:link3d}
% \smallskip
% caption can go here...
% } 
\end{figure}

@@

The NES has a modest CPU that executes game logic. It's a little 8-bit
baby puppy with access to 2 kilobytes of RAM. If you think about it,
the NES's 256x240 pixel screen is already 61,440 pixels, which if it
used the entire 2k of RAM to represent it, we'd have only
$0.2\overline{6}$ bits of RAM per pixel, which makes no sense. At
1.7~MhZ, and a generous 1 instruction per pixel, we'd only be able to
render 28 frames per second! This doesn't make sense, and this is
because the CPU does not render the graphics on a NES.

The NES also contains a custom Picture Processing Unit, or PPU, the
8-bit baby puppy's baby ppupy brother. This thing outputs 60 frames of
NTSC video per second. In fact, since the PPU also drives a CPU
interrupt for each scanline and vblank, and most games use
this for timing,%
%
\footnote{Fun fact: NTSC (``Never Twice the Same Color''), the video
  standard used in the Americas, has a $\sim\!\!60$ Hz frame rate. But
  PAL (``Pointless and Lousy''), used in Europe and Asia, has a $\sim\!\!50$
  Hz refresh rate. This means that many games run 16\% slower in
  Europe than the US, so if you grew up playing there, you were
  playing on {\em Easy} mode. {\tt\ ;-)}}
%
in many ways the CPU is subservient to the PPU, kind of like the PPU
is the {\it true} pure heart of the NES. In the context of this paper, you
could say that the CPU is the Ocarina of Time and the PPU is the
Legend of Zelda. It's just an analogy. The PPU is very complicated and
has tetrahedra for hands, like clubs. The CPU communicates with
the PPU not by drawing pixels, but by giving a high-level description
of the screen (by writing it into PPU memory). The chief insight of
this work is that the PPU representation can be interpreted and
rendered in 3D instead of 2D. This allows us to use the exact original
game logic and only change the way it's viewed. Sometimes this even
allows us to see aspects of the game that are actually present but
were not visible on NES hardware. Often it makes the game more
difficult by making things invisible (enemies behind the player or
behind walls) and inducing motion sickness.

@@

@@

\begin{tabular}{rrl}
    & bit \# & bit deets \\
\hline
MSB & 7      & Vertical flip \\
    & 6      & Horizontal flip \\
    & 5      & something weird \\
    & 4      & I don't know \\
    & 3      & unused? who can say \\
    & 2      & bit 2; could be anything \\
LSB & 1,0    & Palette index \\
\end{tabular}

@@

\section[Three to D---$u$ and $v$]{Three to D---$u$ and $v$\footnote{Spoiler alert because I think
this joke may be too abstruse. ``Free to Be You and Me'' was a
semi-famous kids record (here meaning actual vinyl) by 
Marlo~Thomas. D is ``dimensions.'' $u$ and $v$ are common names of
the axes for texture coordinates in computer graphics.}}

@@

\paragraph{Extracting sprites.}
Sprites are the most difficult. Aside from just being more fiddly to
implement, it is tricky to recover an important high-level structure.
Although we know the location of each individual sprite and the two tiles
(Zelda uses ``tall' 8x16 sprites; see Section~\ref{sec:sprites}) that
make it up, most game objects (like enemies) are in fact drawn with
{\it two} sprites. These sprites could be in any of the 64 sprite
slots (it's not at all obvious how Zelda allocates these, or that it
would be similar for other games). Adding to the complexity, sprites
like Link's sword or the pulsing balls of plasma that frogmen shoot
at you from the water {\it just for being an elf} are just a single
sprite, and bosses can be like 6 or more sprites (Figure~\ref{fig:spider}). If we
promote sprites to 3D without knowing which ones are supposed to be
part of the same object, then the enemies get cut up along their
sprite seems and it looks super wrong.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\linewidth]{spider}
\end{center}\vspace{-0.1in}
\caption{Link's spider nemesis, made of six 8x16 sprites. The sprites
have adjacent edges, so they end up being fused into one merged sprite.
} \label{fig:spider}
\end{figure}

@@

\begin{itemize}
\item Create 64 ``pre-sprites'', which are baby sprites. These know
  their coordinates and have the index of a ``parent'' pre-sprite
  (initialized to $-1$).
\item While filling these in, consult a hash-map to see if there are
  already pre-sprites exactly aligned in each of the cardinal
  directions. If so, use a union-find--like algorithm\cite{tarjan1975efficiency}
  to efficiently update its parent index; each fused sprite has one
  canonical representative whose parent index is $-1$. Add the
  pre-sprite to the hash-map.
\item Another pass pushes the min/max bounds to the canonical representative,
  so that we know how big the fused sprite is.
\item Another pass draws all sprites into the texture data for their
  canonical representative, at the appropriate location. This pass must
  happen in the same order as on the NES hardware, so that overlay effects
  are correct.
\item Finally, create proper sprites to return for each of the fused sprites.
  Copy the used region of the texture data for these to the video card.
\end{itemize}

@@

We also have to face in the camera in the same direction as Link, to
make the perspective first-person. Link's orientation is at memory
location \verb+0x98+.

Link moves smoothly (with pixel resolution) around the screen, but
turns abruptly. This is sickening to watch. I apply some smoothing to
Link's angle, by treating the location from memory as a ``target''
angle and twisting towards it (along the shortest angular path) at a
limited speed.

Knowing what memory locations hold the player's position and
orientation also involves some manual work. You could imagine
heuristics for trying to determine this, by seeing which bytes tend to
go up when pressing right on the controller, and so on. However, it's
easy to find these values manually or to even look them up for popular
games.

@@

\begin{figure*}[t]
\begin{center}
\includegraphics[width=\linewidth]{bridgewide}
\end{center}\vspace{-0.1in}
\caption{A representative view of the Legend of Zelda in glorious
  3D.} \label{fig:bridgewide}
\end{figure*}

Finally, we can render the scene. I used ``classic'' OpenGL interface,
even though it is terrible, for the sake of making good on the
excellent pun that is this paper's title. The final result looks like
Figures~\ref{fig:bridgewide} and~\ref{fig:alone}.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\linewidth]{alone}
\end{center}\vspace{-0.1in}
\caption{The first cave, where this man gives you a sword and then dies,
rendered in 3D. Letters are treated as walls so you can see 'em upright,
but then of course the second line is in front of the first. The 2D NES
screen is superimposed on the top left for comparison.
} \label{fig:alone}
\end{figure}

Rendering with OpenGL is quite straightforward given the geometry
extracted from the previous sections. I try to preserve the pixel
aesthetic by using \verb+GL_TEXTURE_MAG_FILTER+ = \verb+GL_NEAREST+,
but at this point we can use whatever approach to 3D graphics most
titillates the youth. Full-screen multisample anti-aliasing? Sure.
Screen Space Ambient Occlusion? I hear that this creates physically
realistic shadows on the cheap, sure. Stencil buffer shadows?
Volumetric fog? God rays? Depth-of-field? Lens flares? You got it.

\subsection{Quirks}
There are some problems or quirks with the current approach.

\paragraph{Extramural geometry.}
In Zelda, some of the tiles on screen are used to draw a menu at the
top, which is not part of the gameplay area. This still gets
interpreted as part of the geometry of the world, of course
(Figure~\ref{fig:offscreen}). It would be a simple matter to crop the
region of interest (especially if done manually), and it would even be
natural to overlay it on a heads-up display. But I like it over there.
Why shouldn't menus be 3D as well?

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\linewidth]{offscreen}
\end{center}\vspace{-0.1in}
\caption{ The game menu visible to the North because the ocean does
  not get in its way. Did you know that if you were able to see beyond
  the cluttering trees and mountains at the horizon of Earth, you'd
  see your own status menu in real life? } \label{fig:offscreen}
\end{figure}

\paragraph{Positive self-image.}
We place the camera at Link's position and orientation, so that the
player ``becomes'' Link. However, Link is still represented by a
sprite in the game, and we do not treat it specially. This means that
the fused Link billboard is always in the scene geometry right at the
location of the camera. The upshot is that whenever the camera is
facing south, an enormous Link graphic fills the entire screen,
reminding you that he has always been following a few pixels behind
you, haunting you (Figure~\ref{fig:megalink}). This sprite could be
rejected by seeing that it is coincident with the player's location
(which we extract) or by its particular content. We could also set the
``near plane'' so that nearby sprites are not visible. However, the
everpresent looming spectre of a ghost mimicking our every move is a
good reminder of our mortality, so I kept it.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\linewidth]{megalink}
\end{center}\vspace{-0.1in}
\caption{Look who's behind you! WE ARE FRIENDS FOREVER {\tt:-) :-)}
} \label{fig:megalink}
\end{figure}

\paragraph{Surprising sprite fusion.}
Sprite fusion works well for merging sprites that are part of the same
logical object in Zelda. (It may not work as well in other games where
e.g. a monster's legs are moved independently from its body.) However,
it occasionally mistakenly fuses two sprites that just incidentally
share an edge (Figure~\ref{fig:badfusion}). This usually persists for
just a tiny moment since the sprites are moving. But who's to say that
as those monsters high-five, they aren't forming a momentary totem?
Only Shigeru {\it froakin'} Miyamoto can say that isn't what he intended.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\linewidth]{bad-sprite-fusion}
\end{center}\vspace{-0.1in}
\caption{Two Octoroks that were incorrectly fused because they have
constituent sprites that have adjacent edges (the fused sprite shape is
not rectangular). The formation is a Hexadecirok.
} \label{fig:badfusion}
\end{figure}

\paragraph{Texture orientation.} The textures of both backgrounds and sprites
often have strange and inconsistent perspective ideas in video games.
For example, the Octorok points its nose in the direction it's moving,
because it is always viewed ``from above''. However, when Link is
facing south, you see his front (not the top of his head facing down),
and when facing west you see his side. This doesn't make sense, and
makes it impossible to apply any consistent rule for mapping sprites
into world orientation. This mostly just produces humorous results,
like when Link throws his sword and it moves away from him in first
person view: It may be pointing towards the sky, towards the floor, or
to either side, but never in the direction it's thrown. Even if we
were to render the sword ``in plane'', its handle should always be
oriented towards the camera (which would not be the case). No single
policy will work for all cases, but perhaps heuristics can look for
the use of the horizontal and vertical flip bits to figure out what
sprites are rotatable, and then rotate them relative to the camera.

\section{Future work}

This work is literally in progress, and due to Draconian SIGBOVIK
deadlines, this paper is figuratively already out of date by the time
you read it. You can check out the project's home page

\medskip
\begin{center}
\url{http://tom7.org/zelda}
\end{center}
\medskip

and maybe download it and play it, and probably watch a video
demonstration. There may be some new features. Depends what I
get to, you know?

\subsection{Control}
Currently, the game is viewed first person but played with native
controls; this means that pressing ``up'' always moves link to the
North, turning the camera if necessary. This is super weird. However,
it is pretty easy to remap first person controls into native controls,
as long as there is a way of implementing that intent (for example, it
is not possible to ``strafe'' in Zelda). We can change the camera
angle arbitrary, implementing ``mouselook'' or virtual reality
support. Since the z axis is otherwise unused, we could allow the player
to jump, though this would do nothing in the game itself.

\subsection{Other games}
This program was built in such a way that it should take minimal
effort make other NES games 3D, perhaps with humorous results. There
are some issues that need to be dealt with. A common type of game is
the side-scrolling platformer, like Super Mario Bros.. At least two
important issues are not yet handled: Mario has a different coordinate
system than Zelda (it is ``side view'' rather than ``top down''), so
we need to support that rendering style and control/camera scheme. No
big deal. Second, this game scrolls the screen rather than using
single-screen layouts; this uses the PPU in a different and somewhat
more complex way. Many games will use advanced PPU tricks that cannot
be interpreted in a generic high-level way.

\subsection{A Link to the cast} \label{sec:cast}

One dissatisfying problem with the current approach is that it
requires some manual effort to map different tile types to geometry.
One could imagine various heuristics (even a image classifier) to
provide defaults or to remove the manual input entirely. An exciting
angle is to use the game's own understanding of these tiles. The main
thing that differentiates a ``wall'' from a ``floor'' is whether the
player can walk over it. Clearly somewhere in the game's code it
``knows'' what is floor and what is wall. We could extract that
directly, but this would require a different manual step.
Alternatively, we could determine it experimentally. A simple
heuristic would keep track of what tiles the player has walked on. But
we are not limited to what the player actually does; since we are
emulating, we can also emulate counterfactual scenarios. Imagine that
as we draw a first person view, we cast lines out from the camera
towards the geometry, as in a ray tracer. Except that instead of
tracing light rays until an intersection with scene geometry, we trace
a hypothetical player path by emulating the player moving in that
direction, until Link can move no more. This only requires
understanding the bytes that correspond to the player's location
(which we already need) and how the controller works.\footnote{Most
  rays are diagonal. We can implement a north-northwest diagonal by
  repeatedly moving two pixels north, then one west. We could also set
  the player's position directly to a putative destination and then
  move in cardinal directions to test solidness.} Testing contact with
sprites might give us useful heuristics about how they should be
rendered as well.

There are many ways we can imagine this going wrong, but probably
spectacularly so. Interestingly, this would allow us to render parts
of the game that are not even visible in the PPU yet, since emulating
the player's path ahead far enough would cause the game to switch
to future scenes.

\section{Conclusion}

In this paper I made Zelda three-dimensional so that it can be enjoyed
by the kids. I am fully aware that these ``kids'' are actually also
full-grown adults who themselves resent a slightly younger generation
who inexplicably think of Legend of Zelda Twilight Princess HD as the
canonical episode; these people should get to work at enhancing Ocarina
of Time to add Amiibo support. While you're at it, get rid of the
really long section where you have to be a dog. I didn't like that part.
